Predicting the Outcome of Patient-Provider Communication Sequences using Recurrent Neural Networks and Probabilistic Models

### 2018-01-10 paper-amia-summit-2018/amia_summit_2018.tex
camera ready version

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered observation sequences generated by a physiological, genomic or psychological process to make predictions about the outcome of that process arises in
many domains of clinical informatics. In this paper, we focus on the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviewing.
Specifically, we formulate the problem of predicting the success of a motivational interview (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of
coded patient-provider communication exchanges as binary sequence classification.
%Severe imbalance between positive and negative training sequences is an additional challenge that arises in the context of motivational interviewing.
We propose two methods that are based on deep learning, such as Recurrent Neural Networks (RNNs), and probabilistic models, such as Markov Chain (MC) and Hidden Markov Model (HMM), and compare the
accuracy of these methods using communication sequences annotated with behavior codes from real-life motivational interviews. Our experiments indicate that the deep learning-based method is
significantly more accurate than the methods based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM,
respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate
that the proposed method can be used for real-time monitoring of the progression of clinical interviews and identification of the most effective communication strategies, which significantly decreases
the effort required to develop psychological interventions and increases their effectiveness.}

\section*{Introduction}
Temporally ordered sequences of discrete or continuous observations generated by genomic, psychological or psychological processes arise in many different domains of clinical informatics (e.g.,
DNA sequences, gene activations, ECG measurements, laboratory results, diagnostic codes, utterances during clinical interviews). Classification (or categorization) is a type of analysis of those
sequences that has a broad range of important practical applications, from protein classification \cite{yakhnenko2005discriminatively} and function prediction \cite{deshpande2002evaluation} to
detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as their temporal order, makes sequence classification a
more challenging task than traditional classification. Predicting the outcome of those sequences (e.g. physiological or behavior response) can also be formulated as sequence classification.

In general, sequence classification methods fall into one of the three major categories: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature
vector and apply a standard supervised learning methods, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision.
Feature-based methods for sequence classification have had limited success since traditional bag-of-words feature representation methods cannot easily account for the order of observations in a
sequence.
%Shapelet \cite{ye2009time} and pattern based techniques \cite{kudenko1998feature, lesh1999mining} as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of
%standard classifiers as well.
Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric. The most commonly used distance metric is
Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}. However, distance metrics are primarily designed for time series data, in which the observations are discretized by time points.
The third type of sequence classification methods first creates a probabilistic model, such as Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class
using the training data and then classifies new sequences by applying the constructed models. MC and HMM explicitly capture short-term temporal dependencies (e.g. first- and second-order MC and HMM
\cite{kundu1988recognition}) between adjacent observations in a sequence. By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short-
and long-term dependencies and were shown to be effective for modeling different types of sequential data. Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which
successfully addresses the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and
handwriting \cite{nion2013handwritten} recognition to health informatics \cite{lipton2015learning, choi2016doctor}. Specifically, LSTM was used as part of a multi-label classification method to
recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}. LSTM was also effectively used for predicting
the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}. A further simplification and improvement of LSTM model, called the Gated Recurrent
Unit\cite{chung2014empirical} (GRU), was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.

In this paper, we address the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of
Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that
approximately one-third (31.8\%) of US children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and
have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI),
an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires,
ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior
change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of
patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since
manual examination and sequential pattern analysis of transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years.
Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the
mechanisms of MI's efficacy, in particular.

Our goal is to test the applicability of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for predicting the success of clinical interviews (i.e. eliciting a
particular type of patient behavioral response, such as CHT) and dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on a sequence of
coded previous patient-provider communication exchanges in the same interview. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored
several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook
\cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability
of state-of-the-art methods for sequence modeling to analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavior in a context of motivational
interviews, in particular.

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists
of an MI interview session involving counselor, adolescent, and caregiver. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, in which the codes are grouped into
patient (adolescent and caregiver) codes and counselor codes. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in
positive change talk (CHT+) and commitment language (CML+) statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk (CHT-) or commitment language (CML-) and the
sequences, in which no change talk or commitment language statements occur. Out of 5143 observed sequences, 4225 were positive and 918 were negative. Successful sequences had an average length of
9.79 utterances, while unsuccessful sequences had on average 9.65 utterances. For each of the probabilistic models (MC and HMM), two models were trained, one model was trained using successful
sequences and another one was trained using unsuccessful sequences.

%Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}. \\
%
%\begin{table}[h]
%\centering
%\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
%\label{tab:data_dist}
%  \begin{tabular}{|l|l|l|l|l|}
%  \hline
%   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
%Successful sequences & 4225 & 82.15\% & 9.79 \\\hline
%Unsuccessful sequences & 918 & 17.85\% & 9.65 \\\hline
%  \end{tabular}
%\end{table}

Our dataset is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the
performance of an employed algorithm for an imbalanced data, since simply predicting the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to
handle class imbalance properly. Many solutions have been proposed to deal with the imbalanced datasets, which can be divided into two categories: data level and algorithmic level. In this study, we
utilized under-sampling and over-sampling for balancing the samples in different classes in our dataset. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used
method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between
minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the
clusters of samples by the cluster centroid in a $k$-means clustering algorithm.

A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. Annotation column shows successful sequences of behavior codes from top to bottom. For example,
($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive change talk at the end. As
can be seen, similar utterances, such as  ``Yeah" and ``Yes'', can have different behavior codes (CHT+ and HUPW, respectively) depending on the context.\\

\begin{table}[h]
\caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}    
\label{tab:anno_examp}
\centering
\begin{tabular}{|l|p{3.2cm}|l|p{8cm}|}
\hline
\bf{Annotation}  & \bf{Behavior} & \bf{Speaker} & \bf{Text} \\\hline
SS & Structure Session & Counselor & Okay. Can I meet with Xxxx alone for a few minutes? \\\hline
OQO & Open-ended question, other & Counselor & So, Xxxx, how you doing? \\\hline
HUPO &    High uptake, other    & Adolescent &    Fine \\\hline
OQTBN &    Open-ended question, target behavior neutral & Counselor &    That's good.  So, tell me  how do you feel about your weight? \\\hline
CHT+ &    Change talk positive    & Adolescent &    It's not the best. \\\hline
CQECHT+ & Closed question, elicit change talk positive & Counselor & It's not the best? \\\hline
CHT+ &    Change talk positive &    Adolescent & Yeah \\\hline
CQTBN &    Closed question, target behavior neutral  & Counselor &    Okay, so have you tried to lose weight before? \\\hline
HUPW &    High uptake, weight & Adolescent &    Yes \\\hline
\end{tabular}
\end{table}

>>>>

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered sequences of observations generated by a molecular, physiological or psychological process to make predictions about the outcome of that process arises in many domains of clinical informatics. In this paper, we focus on predicting the outcome of patient-provider communication sequences in the context of clinical dialog. Specifically, we consider prediction of the motivational interview success (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of coded patient-provider communication exchanges as a sequence classification problem. We propose two solutions to this problem, one that is based on Recurrent Neural Networks (RNNs) and another that is based on Markov Chain (MC) and Hidden Markov Model (HMM), and compare the accuracy of these solutions using communication sequences annotated with behavior codes from the real-life motivational interviews. Our experiments indicate that the deep learning-based approach is significantly more accurate than the approach based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM, respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate that the proposed method can be used for real-time monitoring of clinical interviews progression and more efficient identification of effective provider communication strategies, which can significantly decrease the effort required to develop behavioral interventions and increases their effectiveness.}

\section*{Introduction}
Temporally ordered sequences of discrete or continuous observations generated by molecular, psychological or psychological process(es) arise in many different areas of biology and medicine (e.g., DNA base-pairs, protein sequences, ECG measurements, laboratory results, diagnostic codes, utterances  in clinical dialog). Classification (or categorization) is a type of analysis of those sequences that has a broad range of important practical applications, from protein function \cite{yakhnenko2005discriminatively} or structure \cite{deshpande2002evaluation} prediction to detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as temporal order and potential dependencies between observations, makes sequence classification a more challenging task than classification of independent observations. Predicting the outcome of those sequences (e.g. physiological or behavioral response) can also be viewed as a sequence classification problem.

In general, sequence classification methods fall into one of the three major classes: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature vector and apply a standard supervised machine learning method, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision. The methods in this class have had limited success, since traditional feature representation methods cannot easily account for the order of and dependencies between observations in a
sequence.

Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric. The most commonly used distance metric is Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}. However, distance metrics are primarily designed for time series data, in which the observations are discretized by timestamps. The third type of sequence classification methods first creates a probabilistic model, such as the Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class based on the training data and then classifies new sequences by applying the created models. While MCs and HMMs can capture first- and second-order dependencies between adjacent observations in a sequence, learning higher-order dependencies with these models requires prohibitively large amounts of data. By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short- and long-term dependencies and were shown to be effective at modeling different types of sequential data \cite{lipton2015critical}. Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which successfully addressed the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and handwriting recognition\cite{nion2013handwritten} to health informatics \cite{lipton2015learning, choi2016doctor}. Specifically, LSTM was used as part of a multi-label classification method to recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}. LSTM was also effectively used for predicting the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}. A further simplification and improvement of LSTM model, called the Gated Recurrent
Unit (GRU)\cite{chung2014empirical}, was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.

In this paper, we address the problem of predicting the outcome of coded patient-provider communication sequences in the context of clinical dialog. Specifically, we focus on predicting the success (i.e. eliciting a particular type of patient behavioral response) of motivational interviews with obese adolescents and their caregivers based on an observed sequence of coded patient-provider communication exchanges during those interviews. Childhood obesity is a serious public health concern in the United States. Recent estimates indicate that approximately one-third (31.8\%) of U.S. children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, collectively referred to as ``change talk'' (CHT), consistently predict the actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of providers to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and analysis of MI interview transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective behavioral interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

Our goal is to compare the accuracy of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for the task of predicting the success of clinical interviews (i.e. eliciting a particular type of patient behavioral response, such as CHT) at any point during a clinical interview based on a sequence of coded previous patient-provider communication exchanges in the same interview, which we consider as a sequence classification problem. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook \cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability of state-of-the-art methods for sequence modeling to the analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.

\section*{Methods}
\subsection*{\textit{Data collection}}

The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which consist of a total of 50,239 segmented and annotated utterances. Each transcript consists of an MI interview session, which typically involves a counselor, an adolescent and a caregiver. The utterances were annotated based on the MYSCOPE codebook \cite{carcone2013provider}, in which the behavior codes are grouped into the patient (adolescent and caregiver) codes and the counselor codes. Annotated utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences are the ones, which resulted in positive change talk (CHT+) or commitment language (CML+) statements by an adolescent or a caregiver, while unsuccessful sequences are the ones, which resulted in negative change talk (CHT-) or commitment language (CML-), or the ones, in which no change talk or commitment language statements were made.

A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. In this example, $SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$ is a successful sequence, in which a counselor starts with an open-ended question and ultimately is able to elicit a positive change talk statement. As follows from this example, similar utterances, such as ``Yeah'' and ``Yes'', can be assigned different behavior codes (CHT+ and HUPW, respectively), depending on the context.\\

\begin{table}[h]
\caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent. MYSCOPE codes assigned to the utterances and their meaning are shown in the first two columns.}    
\label{tab:anno_examp}
\centering
\begin{tabular}{|l|p{3.2cm}|l|p{8cm}|}
\hline
\bf{Code}  & \bf{Behavior} & \bf{Speaker} & \bf{Utterance} \\\hline
SS & Structure Session & Counselor & Okay. Can I meet with Xxxx alone for a few minutes? \\\hline
OQO & Open-ended question, other & Counselor & So, Xxxx, how you doing? \\\hline
HUPO &    High uptake, other    & Adolescent &    Fine \\\hline
OQTBN &    Open-ended question, target behavior neutral & Counselor &    That's good.  So, tell me  how do you feel about your weight? \\\hline
CHT+ &    Change talk positive    & Adolescent &    It's not the best. \\\hline
CQECHT+ & Closed question, elicit change talk positive & Counselor & It's not the best? \\\hline
CHT+ &    Change talk positive &    Adolescent & Yeah \\\hline
CQTBN &    Closed question, target behavior neutral  & Counselor &    Okay, so have you tried to lose weight before? \\\hline
HUPW &    High uptake, weight & Adolescent &    Yes \\\hline
\end{tabular}
\end{table}

The resulting experimental dataset was highly imbalanced. Out of 5143 observed sequences, 4225 or 82.15\% were positive and only 918 or 17.85\% were negative. No major differences were observed in the average length of successful (9.79 utterances) and unsuccessful (9.65 utterances) sequences.

Since severely imbalanced datasets often distort the true performance of a classification method relative to a simple ``majority vote'' baseline (e.g. simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset). Therefore, it is important to
properly address class imbalance. We evaluated the performance of probabilistic and deep learning methods using both under-sampling and over-sampling for balancing the number of samples in different classes. Synthetic Minority Over Sampling Technique (SMOTE) \cite{chawla2002smote} is a widely used oversampling method for imbalanced datasets, in which new synthetic examples are generated for minority classes. Specifically, we generated synthetic examples at the borderline between the majority and minority classes \cite{nguyen2011borderline}. On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the clusters of samples identified by the $k$-means clustering algorithm with the cluster centroids.

Two MC and HMM models were trained, one model was trained on successful sequences, while another was trained on unsuccessful sequences.
